Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-05-26 paper/graph.tex
Added scalability discussion
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@620 3fc257bb-8550-0410-81b3-a5c103da83f7

The algorithm in Figure~\ref{fig:pr} performs these
iterations.  The nonzeros of matrix $A$ are initialized as described above;
the PageRank vector $x$ is initialized uniformly.  A MapReduce object $MT$
contains the indices of all-zero rows of $A$, corresponding to vertices
with no outlinks.  In step 1 of a PageRank interation, $MT$ is used 
in dot products to compute adjustments to the
PageRank vector to represent uniform probability of going from a leaf page
to any other page.  Adjustments for random jumps from any page are also
computed in this step.  Step 2, the matrix-vector product $A^T x$, is the most
expensive part of the computation, requiring two passes over the nonzero
structure of $A$.  In the first pass, {\it collate()} 
gathers all row entries $a_{ij}$ with their
associated $x_i$ entry, and {\it reduce()} computes $a_{ij} x_i$.  
In the second pass, 
{\it collate()} gathers, for each $j$, all contributions to the
column sum $\sum_i a_{ij} x_i$, which is computed by a second {\it reduce()}. 
Steps 3 and 4 adjust and scale the 
product vector; the work is proportional to $|V|$.  Step 6 computes the
residual to determine whether the PageRank vector has converged; work is
again proportional to $|V|$.  Throughout the algorithm, we exploit the
ability to call MPI\_Allreduce to compute global norms and residuals.

%Input from R-MAT generator: \\
%\> MapReduce object $V$ containing all graph vertices:  Key=$V_i$, Value = NULL.  \\
%\> MapReduce object $E$ containing all graph edges: Key=$V_i$, Value = $V_j$. \\
%Initialize MapReduce object for matrix $A$  and \\
%MapReduce object $MT$ containing all vertices with no outlinks.\\
%\> Copy: $A = E$ \\
%\> Add: $V$ to $A$ \\
%\> Collate $A$: Vertex is Key \\
%\> Reduce $A$:  \\
%\> \> Input:  Key = $V_i$; MultiValue = NULL,$V_j$,$V_k$,... \\
%\> \> If (${nvalues} > 1$) \\
%\> \> \> Compute: $d_{out}(V_i) = {nvalues} - 1$. \\
%\> \> \> Output:  For each $V_j$ in MultiValue, Key = $V_i$; Value = ($V_j$, $\alpha/d_{out}(V_i)$). \\
%\> \> else \\
%\> \> \> Output to $MT$: Key = $V_i$; Value = 0. \\
%Initialize MapReduce object for pagerank vector $x$. \\
%\> Map:   \\
%\> \> Input:  Graph vertices $V$. \\
%\> \> Output:  For each vertex, Key = $V_i$; Value = $1/|V|$. \\
Input: \\
\> MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
\> MapReduce object $MT$ containing indices of all-zero rows of $A$:  Key = $i$, Value = 0. \\
\> MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> 1. Compute contribution for random jumps and zero-outdegree vertices. \\
\> \> Add: $x$ to $MT$. \\
\> \> Collate $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
\> \> \> Compute:  Local contributions $c_{local}$. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
\> \> MPI\_Allreduce $c_{local}$ with MPI\_SUM to compute global contribution $c_{global}$. \\
\> 2. Compute $y = A^T x$.  \\
\> \> Copy $x$ to $y$. \\
\> \> Add $A$ to $y$. \\
\> \> Collate $y$:  Key is row index of $A$. \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
\> \> Collate $y$:  Key is column index of $A$. \\
\> \> Reduce $y$:  \\
\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
\> 3. Add contribution $c_{global}$ to $y$; compute local max norm $n_{local}$ of $y$. \\
\> \> Map $y$:  \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Compute: $y_i = y_i + c_{global}$; if $y_i > n_{local}$, $n_{local} = y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i$. \\
\> \> MPI\_Allreduce $n_{local}$ with MPI\_MAX to compute global max norm $n_{global}$. \\
\> 4. Scale $y$ by $n_{global}$. \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i / {n_{global}}$. \\
\> 5. Compute residual. \\
\> \> Add $y$ to $x$. \\
\> \> Collate $x$:  Key is index $i$. \\
\> \> Reduce $x$:   \\
\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
\> 6. $x = y$; \\
Output:  PageRank vector $x$.
  \end{tabbing}
 \end{minipage}}\end{center}

 \caption{MapReduce algorithm for PageRank vertex ranking.}

 \label{fig:pr}
\end{figure}

The entire PageRank algorithm can be made more efficient by calling
{\it aggregate()} on the MapReduce objects representing the matrix $A$, 
the all-zero matrix rows $MT$, and the vector $x$ once before the PageRank 
iterations begin.  Since these three MapReduce
objects have the same key types, pre-aggregating the objects moves all
Key-Value pairs with a given key to a single processor.  Thus, many of
the PageRank operations become local operations, replacing 
the {\it collate()} operation (which is equivalent to {\it aggregate()} 
followed by {\it convert()}) with only a local {\it convert()} call. 
Figure~\ref{fig:pr2} shows the
enhanced algorithm; modified operations are marked with an asterisk.
This strategy of pre-aggregating to improve data locality is useful in
many MapReduce algorithms, but assumes that keys are always mapped to
the same processor.

@@

 \caption{Enhanced MapReduce algorithm for PageRank vertex ranking; pre-aggregating the MapReduce objects allows many communication intensive {\it collate()} 
operations to be replaced by with strictly local {\it convert()} operations (as marked by the asterisks).}

>>>>

The algorithm in Figure~\ref{fig:pr2} performs these
iterations.  The nonzeros of matrix $A$ are initialized as described above;
the PageRank vector $x$ is initialized uniformly.  A MapReduce object $MT$
contains the indices of all-zero rows of $A$, corresponding to vertices
with no outlinks.  
The PageRank algorithm is made more efficient by calling
{\it aggregate()} on the MapReduce objects representing the matrix $A$, 
the all-zero matrix rows $MT$, and the vector $x$ once before the PageRank 
iterations begin.  Since these three MapReduce
objects have the same key types, pre-aggregating the objects moves all
Key-Value pairs with a given key to a single processor.  Thus, many of
the PageRank operations become local operations, replacing 
{\it collate()} operations (which is equivalent to {\it aggregate()} 
followed by {\it convert()}) with only local {\it convert()} calls. 
In Figure~\ref{fig:pr2}, steps where {\it convert()} operations can be
substituted for {\it convert()} operations are marked with an asterisk.
This strategy of pre-aggregating to improve data locality is useful in
many MapReduce algorithms, but assumes that keys are always mapped to
the same processor.

In step 1 of a PageRank interation, $MT$ is used 
in dot products to compute adjustments to the
PageRank vector to represent uniform probability of going from a leaf page
to any other page.  Adjustments for random jumps from any page are also
computed in this step.  Step 2, the matrix-vector product $A^T x$, is the most
expensive part of the computation, requiring two passes over the nonzero
structure of $A$.  In the first pass, {\it convert()} 
gathers all local row entries $a_{ij}$ with their
associated $x_i$ entry, and {\it reduce()} computes $a_{ij} x_i$.  
In the second pass, 
{\it collate()} gathers, for each $j$, all contributions to the
column sum $\sum_i a_{ij} x_i$, which is computed by a second {\it reduce()}. 
Steps 3 and 4 adjust and scale the 
product vector; the work is proportional to $|V|$.  Step 6 computes the
residual to determine whether the PageRank vector has converged; work is
again proportional to $|V|$.  Throughout the algorithm, we exploit the
ability to call MPI\_Allreduce to compute global norms and residuals.

%\begin{figure}[htb]
% \begin{center}\fbox{\begin{minipage}{\textwidth} \begin{tabbing}
%  xxxx\=xxxxx\=xxxxx\=xxxxx\=xxxxxxxxxxxxxxxxxxxxxxxxxxx\= \kill
%Input: \\
%\> MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
%\> MapReduce object $MT$ containing indices of all-zero rows of $A$:  Key = $i$, Value = 0. \\
%\> MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
%While (${residual} > {tolerance}$) \\
%\> 1. Compute contribution for random jumps and zero-outdegree vertices. \\
%\> \> Add: $x$ to $MT$. \\
%\> \> Collate $MT$:  Vertex is Key. \\
%\> \> Reduce $MT$:   \\
%\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
%\> \> \> Compute:  Local contributions $c_{local}$. \\
%\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
%\> \> MPI\_Allreduce $c_{local}$ with MPI\_SUM to compute global contribution $c_{global}$. \\
%\> 2. Compute $y = A^T x$.  \\
%\> \> Copy $x$ to $y$. \\
%\> \> Add $A$ to $y$. \\
%\> \> Collate $y$:  Key is row index of $A$. \\
%\> \> Reduce $y$: \\
%\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
%\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
%\> \> Collate $y$:  Key is column index of $A$. \\
%\> \> Reduce $y$:  \\
%\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
%\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
%\> 3. Add contribution $c_{global}$ to $y$; compute local max norm $n_{local}$ of $y$. \\
%\> \> Map $y$:  \\
%\> \> \> Input:  Key = $i$, Value = $y_i$. \\
%\> \> \> Compute: $y_i = y_i + c_{global}$; if $y_i > n_{local}$, $n_{local} = y_i$. \\
%\> \> \> Output:  Key = $i$, Value = $y_i$. \\
%\> \> MPI\_Allreduce $n_{local}$ with MPI\_MAX to compute global max norm $n_{global}$. \\
%\> 4. Scale $y$ by $n_{global}$. \\
%\> \> Map $y$: \\
%\> \> \> Input:  Key = $i$, Value = $y_i$. \\
%\> \> \> Output:  Key = $i$, Value = $y_i / {n_{global}}$. \\
%\> 5. Compute residual. \\
%\> \> Add $y$ to $x$. \\
%\> \> Collate $x$:  Key is index $i$. \\
%\> \> Reduce $x$:   \\
%\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
%\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
%\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
%\> 6. $x = y$; \\
%Output:  PageRank vector $x$.
  %\end{tabbing}
 %\end{minipage}}\end{center}
%
 %\caption{MapReduce algorithm for PageRank vertex ranking.}
%
 %\label{fig:pr}
%\end{figure}

@@

 \caption{MapReduce algorithm for PageRank vertex ranking; pre-aggregating the MapReduce objects allows many communication intensive {\it collate()} 
operations to be replaced by with strictly local {\it convert()} operations (as marked by the asterisks).}

### 2010-05-26 paper/results.tex
Added scalability discussion
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@620 3fc257bb-8550-0410-81b3-a5c103da83f7

\subsection{Comparisons to Other Implementations}

@@

We ran each of the algorithms on four R-MAT graphs of
different sizes, each on a varying number of processors.  Details of the
input data are shown in Table~\ref{t:rmats}.
The {\it tiny} problem size (around 8M edges) can typically be used
on a single processor without significant out-of-core operations.
The {\it small} problem size (around 134M edges) is a 
problem that can be run
on a single processor with out-of-core operations; larger processor 
configurations, however, do not necessary need out-of-core operations.
The {\it medium} problem size (around 2B
edges) can be used on a few processors.  The {\it large} problem size
(around 34B edges) requires most of the machine to run.
All data sets used R-MAT parameters $(a, b, c, d) = (0.57, 0.19, 0.19, 0.05)$.

\begin{table}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Data & \# of    & \# of & Maximum \\
Set  & vertices & edges & vertex degree\\
\hline
tiny   &$2^{20} \approx 1M$ & $2^{23} \approx 8M$ &   \\
small  &$2^{24} \approx 17M$ & $2^{27} \approx 134M$ &   \\
medium &$2^{28} \approx 268M$ & $2^{31} \approx 2B$&   \\
large  &$2^{32} \approx 4B$ & $2^{35} \approx 34B$ &   \\
\hline
\end{tabular}
\caption{Characteristics of R-MAT input data for graph algorithm
experiments.}
\label{t:rmats}
\end{table}

@@

Discuss R-MAT generation times.

Discuss PageRank performance.  Include Trilinos and other
algorithms and machines?

Discuss triangle finding times.

Discuss connected component identification times.

Discuss Luby maximally independent sets times.

Discuss SSSP times.

Finally, we compare our MapReduce implementations with a multi-threaded
implementations
in the Multi-Threaded Graph Library (MTGL)~\cite{MTGL}, and
distributed-memory matrix-based implementations using the linear
algebra library Trilinos~\cite{Trilinos-Overview}.
In the MTGL~\cite{MTGL} implementation,
rank propagates via adjacency list traversal
in a compressed sparse-row data structure.
To maintain scalability, code must
be written so that a single thread spawns the loop that processes all
in-neighbors of a given vertex; this detail enables the compiler to generate
hotspot-free code.
The matrix-based distributed-memory implementation of PageRank
uses Trilinos' Epetra matrix/vector classes to represent the graph and
PageRank vector.
Rows of matrix $A$ and the associated entries of the PageRank vector $x$
are uniquely assigned to processors; a random permutation of the input
matrix achieves processor load balance.
Interprocessor communication gathers $x$ values for matrix-vector
multiplication and sums partial products into the $y$ vector.
Most communication is point-to-point communication,
but some global communication is needed for computing
residuals and norms of $x$ and $y$.

\begin{table}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Data & R-MAT  & R-MAT  & R-MAT  & R-MAT  & \# of    & \# of & Maximum \\
Set  & a      & b      & c      & d      & vertices & edges & vertex degree\\
\hline
nice  & 0.45 & 0.15 & 0.15 & 0.25 & $2^{25}$ & $2^{28}$ & 1108 \\
nasty & 0.57 & 0.19 & 0.19 & 0.05 & $2^{25}$ & $2^{28}$ & 230,207\\
\hline
\end{tabular}
\caption{Characteristics of R-MAT input data for PageRank and Connected
Components experiments.}
\label{t:rmat}
\end{table}

We performed our experiments using the R-MAT data sets described in 
Table~\ref{t:rmat}.  
In Figure~\ref{f:prbig}, we show the performance 
of the various PageRank
implementations on distributed memory and multi-threaded architectures.
The MR-MPI implementation was run on the Redstorm parallel computer;
for these experiments,
the data set fit in the memory of Redstorm, so out-of-core operations
were not used.
The MR-MPI implementation demonstrated good scalability up to 1024 processors; 
however, it required an order-of-magnitude
more execution time than the matrix-based implementations on Redstorm.  
This result is likely due to higher volume of
communication in the MapReduce implementation (where all edges are communicated
in each iteration), compared to the Trilinos implementation (where only
vertex-based data are communicated).  The distributed memory matrix-based
implementations are competitive with the multi-threaded implementation
in MTGL on the Cray XMT.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{fig_pagerank.pdf}
\caption{Comparison of PageRank implementations using MapReduce,
Trilinos, and MTGL on R-MAT data sets.}
\label{f:prbig}
\end{figure}

>>>>

@@

We ran each of the algorithms on four R-MAT graphs of
different sizes, each on a varying number of processors.  Details of the
input data are shown in Table~\ref{t:rmats}.
The {\it small} problem size (around 8M edges) can typically be used
on a single processor without significant out-of-core operations.
The {\it medium} problem size (around 134M edges) is a 
problem that can be run
on a single processor with out-of-core operations; larger processor 
configurations, however, do not necessarily need out-of-core operations.
The {\it large} problem size (around 2B
edges) requires out-of-core operations and higher processor counts.
%The {\it x-large} problem size
%(around 34B edges) requires most of the machine to run.
All data sets used R-MAT parameters $(a, b, c, d) = (0.57, 0.19, 0.19, 0.05)$.

\begin{table}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Data & \# of    & \# of & Maximum \\
Set  & vertices & edges & vertex degree\\
\hline
small   &$2^{20} \approx 1M$ & $2^{23} \approx 8M$ &   \\
medium  &$2^{24} \approx 17M$ & $2^{27} \approx 134M$ &   \\
large   &$2^{28} \approx 268M$ & $2^{31} \approx 2B$&   \\
%x-large &$2^{32} \approx 4B$ & $2^{35} \approx 34B$ &   \\
\hline
\end{tabular}
\caption{Characteristics of R-MAT input data for graph algorithm
experiments.}
\label{t:rmats}
\end{table}

@@

\subsection{R-MAT generation results}
Discuss R-MAT generation times.

\subsection{PageRank results}

In Figure~\ref{f:pr}, we show the performance of the MR-MPI PageRank algorithm
compared to a
distributed-memory matrix-based implementation using the linear
algebra toolkit Trilinos~\cite{Trilinos-Overview}.
The matrix-based distributed-memory implementation of PageRank
uses Trilinos' Epetra matrix/vector classes to represent the graph and
PageRank vector.
Rows of matrix $A$ and the associated entries of the PageRank vector $x$
are uniquely assigned to processors; a random permutation of the input
matrix achieves processor load balance.
Interprocessor communication gathers $x$ values for matrix-vector
multiplication and sums partial products into the $y$ vector.
Most communication is point-to-point communication,
but some global communication is needed for computing
residuals and norms of $x$ and $y$.

%We compare our MapReduce implementations with a multi-threaded
%implementations
%in the Multi-Threaded Graph Library (MTGL)~\cite{MTGL}.
%In the MTGL~\cite{MTGL} implementation,
%rank propagates via adjacency list traversal
%in a compressed sparse-row data structure.
%To maintain scalability, code must
%be written so that a single thread spawns the loop that processes all
%in-neighbors of a given vertex; this detail enables the compiler to generate
%hotspot-free code.

%This result is likely due to higher volume of
%communication in the MapReduce implementation (where all edges are communicated
%in each iteration), compared to the Trilinos implementation (where only
%vertex-based data are communicated).

\subsection{Triangle finding results}
Discuss triangle finding times.

\subsection{Maximally independent set results}
Discuss Luby maximally independent sets times.

\subsection{Single-source shortest path results}
Discuss SSSP times.

\subsection{Scalability to large numbers of processors}
Finally, we demonstrate the scalability of MR-MPI to large numbers of 
processors.  
The MR-MPI implementation was run on the Redstorm and Thunderbird
parallel computers
Because these systems do not have local disks for each processor, we selected 
a data set and page sizes that fit in their memory, so out-of-core operations
were not used.  For these experiments, we used an R-MAT data set with 
with $2^{25}$ vertices and $2^{28}$ edges, with parameters given in
Table~\ref{t:rmat}.  We ran both PageRank and Connected Components
algorithms.

\begin{table}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Data & R-MAT  & R-MAT  & R-MAT  & R-MAT  & \# of    & \# of & Maximum \\
Set  & a      & b      & c      & d      & vertices & edges & vertex degree\\
\hline
nice  & 0.45 & 0.15 & 0.15 & 0.25 & $2^{25}$ & $2^{28}$ & 1108 \\
nasty & 0.57 & 0.19 & 0.19 & 0.05 & $2^{25}$ & $2^{28}$ & 230,207\\
\hline
\end{tabular}
\caption{Characteristics of R-MAT input data for PageRank and Connected
Components scalability experiments.}
\label{t:rmat}
\end{table}

In Figure~\ref{f:prbig}, we show the performance 
of the various PageRank
implementations on distributed memory and multi-threaded architectures.
The MR-MPI implementation demonstrated good scalability up to 1024 processors; 
however, it required an order-of-magnitude
more execution time than the matrix-based implementations on Redstorm.  
The distributed memory matrix-based
implementations are competitive with the multi-threaded implementation
in MTGL on the Cray XMT.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{fig_pagerank_big.pdf}
\caption{Comparison of PageRank implementations using MapReduce,
Trilinos, and MTGL on R-MAT data sets.}
\label{f:prbig}
\end{figure}

Similar results were obtained for the Connected Components algorithm, as
we show in Figure~\ref{f:ccbig}.  As with PageRank, the MR-MPI implementation
showed good scalability up to 1024 processors, but required significantly
more time than the Giant Connected Components algorithms using MTGL and
Trilinos.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{fig_cc_big.pdf}
\caption{Comparison of Connected Components implementations using MapReduce,
Trilinos, and MTGL on R-MAT data sets.}
\label{f:prbig}
\end{figure}
